---
title: "Data_Cleaning_Assignment"
output: 
  pdf_document:
      latex_engine: xelatex
date: "2026-04-01"
---

# 1. Data set overview
```{r}
#install.packages("readr")
library(readr)

# Reading data sets
netflix_data <- read.csv("Data/netflix_titles.csv")
imdb_data <- read.csv("Data/TMDB_all_movies.csv")

```


```{r data overview}

# initial exploration
head(netflix_data)
str(netflix_data)
summary(netflix_data)

head(imdb_data)
str(imdb_data)
summary(imdb_data)

```

# 2. Variable Cleaning
The title variable was cleaned to improve consistency between the two datasets and enable accurate merging. All titles were converted to lowercase, leading and trailing whitespace was removed, and special characters were stripped. This ensured that titles with minor formatting differences (like  punctuation or capitalization) would match correctly across datasets.
```{r Title Variable Cleaning}

## 2. Variable 1 Cleaning: Cleaning Title and Year 
library(dplyr)
library(stringr)

# cleaning title
netflix_clean <- netflix_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

imdb_clean <- imdb_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

```

The release year variable in the IMDb dataset was originally stored as a character string within the release_date field. We converted into a numeric format by extracting the year component. Observations with missing or unrealistic years (before 1900 or after 2021) were removed to ensure data accuracy and consistency with the Netflix dataset.
```{r Release Year Variable Cleaning}
# cleaning release year
imdb_clean <- imdb_clean %>%
  mutate(release_year = as.numeric(substr(release_date, 1, 4))) %>%
  filter(!is.na(release_year),
         release_year >= 1900,
         release_year <= 2021)

summary(netflix_clean$release_year)
summary(imdb_clean$release_year)

```

Observations with missing IMDb ratings or vote counts were removed, as these values are essential for analysis. Additionally, Movies with fewer than 50 votes were removed to ensure that IMDb ratings reflect a sufficiently large audience consensus. Low vote counts can lead to unstable and biased ratings, which may distort analysis results. By applying this filter, the dataset better represents reliable measures of content quality, allowing for more accurate evaluation of Netflix’s content strategy and the factors influencing audience ratings.
``` {r IMDB Rating and Votes Variable Cleaning}
## 3. Variable 2 Cleaning: Cleaning imdb rating and votes
imdb_clean <- imdb_clean %>%
  filter(!is.na(imdb_rating),
         !is.na(imdb_votes),
         imdb_votes > 50)
        
```

The cleaned Netflix and IMDb datasets were merged using an inner join on the cleaned title and release year variables. This ensured that only movies present in both datasets were retained. Duplicate entries resulting from the merge were removed to maintain a one-to-one mapping of observations.
``` {r Merging the Datasets}
## 4. Variable 3 Cleaning: Merging the two datasets
# We use an inner join to keep only the movies that exist in BOTH datasets
merged_data <- inner_join(netflix_clean, imdb_clean, by = c("title_clean", "release_year"))

# Removing any duplicate entries that may have occurred during the join
merged_data <- merged_data %>% 
  distinct(title_clean, release_year, .keep_all = TRUE)

```

# 3. Before vs After Comparison

This subsection provides a concise comparison of key statistics for the IMDb ratings before and after cleaning. It includes the total number of rows, the count of missing IMDb ratings, and key descriptive statistics like mean, median, and standard deviation. This gives a numerical snapshot of how the cleaning affected the dataset size and rating distribution.
```{r}
# Installing ggplot2 for visualizations if not already installed
# install.packages("ggplot2")
library(ggplot2)
library(dplyr)
library(knitr)

# --- Comparison of IMDB Data Before and After Cleaning ---
comparison_table <- data.frame(
  Metric = c("Rows",
             "Missing IMDB Ratings",
             "Mean IMDB Rating",
             "Median IMDB Rating",
             "SD IMDB Rating"),
  
  Before = c(
    nrow(imdb_data),
    sum(is.na(imdb_data$imdb_rating)),
    mean(imdb_data$imdb_rating, na.rm = TRUE),
    median(imdb_data$imdb_rating, na.rm = TRUE),
    sd(imdb_data$imdb_rating, na.rm = TRUE)
  ),
  
  After = c(
    nrow(merged_data),
    sum(is.na(merged_data$imdb_rating)),
    mean(merged_data$imdb_rating, na.rm = TRUE),
    median(merged_data$imdb_rating, na.rm = TRUE),
    sd(merged_data$imdb_rating, na.rm = TRUE)
  )
)

kable(comparison_table, digits = 2,
      caption = "Comparison of IMDB Data Before and After Cleaning")

```

This part examines the extent of missing data in all variables before and after cleaning. It confirms the improvement in data completeness and highlights the variables where cleaning had the biggest impact.
``` {r}
# --- Missing values overview ---
# Missing values before cleaning
missing_before <- colSums(is.na(imdb_data))

# Missing values after cleaning
missing_after <- colSums(is.na(merged_data))

missing_before
missing_after

```

Here we visually compare the distributions of IMDb ratings before and after cleaning. The boxplot highlights changes in spread and outliers, while the side-by-side histograms show how the frequency distribution of ratings has shifted.
```{r}
# --- Visualizations ---
# Prepare data for plotting
plot_data_before <- data.frame(
  Rating = imdb_data$imdb_rating,
  State = "Before Cleaning"
)

plot_data_after <- data.frame(
  Rating = merged_data$imdb_rating,
  State = "After Cleaning"
)

plot_data <- rbind(
  plot_data_before[!is.na(plot_data_before$Rating), ],
  plot_data_after
)

# --- Boxplot ---
ggplot(plot_data, aes(x = State, y = Rating, fill = State)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "IMDB Rating Distribution Before and After Cleaning",
       x = "Dataset",
       y = "IMDB Rating") +
  theme(legend.position = "none")

# --- Histogram (Side-by-Side) ---
ggplot(plot_data, aes(x = Rating, fill = State)) +
  geom_histogram(bins = 30, color = "black") +
  facet_wrap(~State) +
  theme_minimal() +
  labs(title = "IMDB Ratings Before vs After Cleaning",
       x = "IMDB Rating",
       y = "Count") +
  theme(legend.position = "none")

```

Finally, this subsection examines the distribution of IMDb votes on a logarithmic scale before and after cleaning. This visualization reveals the wide variation in movie popularity and how filtering has reduced low-vote entries.
```{r}
# --- Votes Distribution (Log Scale) ---
plot_votes_before <- data.frame(
  Votes = imdb_data$imdb_votes,
  State = "Before"
)

plot_votes_after <- data.frame(
  Votes = merged_data$imdb_votes,
  State = "After"
)

plot_votes <- rbind(plot_votes_before, plot_votes_after)

ggplot(plot_votes, aes(x = Votes, fill = State)) +
  geom_histogram(bins = 30) +
  scale_x_log10() +
  facet_wrap(~State) +
  theme_minimal()

```

