---
title: "Data_Cleaning_Assignment"
author: "MM"
output: 
  pdf_document:
      latex_engine: xelatex
date: "2026-04-01"
---

# 1. Data set overview
```{r}
#install.packages("readr")
library(readr)

# Reading data sets
netflix_data <- read.csv("Data/netflix_titles.csv")
imdb_data <- read.csv("Data/TMDB_all_movies.csv")

```


```{r data overview}

# initial exploration
#head(netflix_data)
str(netflix_data)
#summary(netflix_data)

#head(imdb_data)
str(imdb_data)
#summary(imdb_data)

```

# 2. Cleaning Process

## 2.1 Title Variable Cleaning
The title variable was cleaned to improve consistency between the two datasets 
and enable accurate merging. All titles were converted to lowercase, 
leading and trailing whitespace was removed, and special characters were stripped. 
This ensured that titles with minor formatting differences 
(e.g. punctuation or capitalization) would match correctly across datasets.

```{r}
library(dplyr)
library(stringr)

netflix_clean <- netflix_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

imdb_clean <- imdb_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

```


## 2.2 Release Year Variable Cleaning
The release year variable in the IMDb dataset was originally stored as a 
character string within the release_date field. We converted into a numeric format 
by extracting the year component. Observations with years outside the scope of our 
study were removed to ensure data accuracy and consistency with the Netflix dataset.


```{r}
imdb_clean <- imdb_clean %>%
  mutate(release_year = as.numeric(substr(release_date, 1, 4))) %>%
  filter(!is.na(release_year),
         release_year >= 1900,
         release_year <= 2021)

range(imdb_clean$release_year)

#summary(netflix_clean$release_year)
#summary(imdb_clean$release_year)

```


## 2.3 IMDB Rating and Votes Variable Cleaning
Observations with missing IMDb ratings or vote counts were removed. 
Additionally, Movies with fewer than 50 votes were removed to ensure that IMDb 
ratings reflect a sufficiently large audience consensus. Low vote counts can lead to 
unstable and biased ratings, which may distort analysis results. 


``` {r}
imdb_clean <- imdb_clean %>%
  filter(!is.na(imdb_rating),
         !is.na(imdb_votes),
         imdb_votes > 50)
        
```

## 2.4 Merging the Datasets
Duplicate entries were first removed to maintain a one-to-one mapping of observations.
The cleaned Netflix and IMDb data sets were merged using an inner join on the 
cleaned title and release year variables. Release year was used due to the 
occurrence of remakes. This process ensured that only movies present 
in both datasets were retained.  


``` {r}
## 4. Variable 3 Cleaning: Merging the two datasets

# Clean both datasets first
netflix_unique <- netflix_clean %>% distinct(title_clean, release_year, .keep_all = TRUE)
imdb_unique <- imdb_clean %>% distinct(title_clean, release_year, .keep_all = TRUE)

# Join the unique data
merged_data <- inner_join(netflix_unique, imdb_unique, by = c("title_clean", "release_year"))

```

# 3. Summarization Before and After Cleaning
Comparison of key statistics for the IMDb ratings before and after cleaning, including 
the total number of rows, the count of missing IMDb ratings, and key descriptive 
statistics like mean, median, and standard deviation. This gives a numerical 
snapshot of how the cleaning affected the dataset size and rating distribution.


```{r}
library(ggplot2)
library(dplyr)
library(knitr)

# --- Comparison of IMDB Data Before and After Cleaning ---
comparison_table <- data.frame(
  Metric = c("Rows",
             "Missing IMDB Ratings",
             "Mean IMDB Rating",
             "Median IMDB Rating",
             "SD IMDB Rating"),
  
  Before = c(
    nrow(imdb_data),
    sum(is.na(imdb_data$imdb_rating)),
    mean(imdb_data$imdb_rating, na.rm = TRUE),
    median(imdb_data$imdb_rating, na.rm = TRUE),
    sd(imdb_data$imdb_rating, na.rm = TRUE)
  ),
  
  After = c(
    nrow(merged_data),
    sum(is.na(merged_data$imdb_rating)),
    mean(merged_data$imdb_rating, na.rm = TRUE),
    median(merged_data$imdb_rating, na.rm = TRUE),
    sd(merged_data$imdb_rating, na.rm = TRUE)
  )
)

kable(comparison_table, digits = 2,
      caption = "Comparison of IMDB Data Before and After Cleaning")

```

This part examines the extent of missing data in all variables before and after 
cleaning, and highlights the variables where cleaning had the biggest impact.

``` {r}
# --- Missing values overview ---

# Missing values before cleaning
missing_before <- colSums(is.na(imdb_data))

# Missing values after cleaning
missing_after <- colSums(is.na(merged_data))

missing_before
missing_after

```


Here we visually compare the distributions of IMDb ratings before and after cleaning. 
The boxplot highlights changes in spread and outliers, while the side-by-side 
histograms show how the frequency distribution of ratings has shifted.


```{r}
# --- Visualizations ---
# Prepare data for plotting
plot_data_before <- data.frame(
  Rating = imdb_data$imdb_rating,
  State = "Before Cleaning"
)

plot_data_after <- data.frame(
  Rating = merged_data$imdb_rating,
  State = "After Cleaning"
)

plot_data <- rbind(
  plot_data_before[!is.na(plot_data_before$Rating), ],
  plot_data_after
)

# --- Boxplot ---
ggplot(plot_data, aes(x = State, y = Rating, fill = State)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "IMDB Rating Distribution Before and After Cleaning",
       x = "Dataset",
       y = "IMDB Rating") +
  theme(legend.position = "none")

# --- Histogram (Side-by-Side) ---
ggplot(plot_data, aes(x = Rating, fill = State)) +
  geom_histogram(bins = 30, color = "black") +
  facet_wrap(~State) +
  theme_minimal() +
  labs(title = "IMDB Ratings Before vs After Cleaning",
       x = "IMDB Rating",
       y = "Count") +
  theme(legend.position = "none")

```

Finally, this subsection examines the distribution of IMDb votes on a 
logarithmic scale before and after cleaning. This visualization reveals the 
wide variation in movie popularity and how filtering has reduced low-vote entries.

```{r distributions}

# Combine and filter 0s/NAs 
plot_votes <- bind_rows(
  mutate(imdb_data, State = "Before"),
  mutate(merged_data, State = "After")
) %>%
  filter(imdb_votes > 0) # Log scale can't handle 0 or negative numbers

# Create plot 
ggplot(plot_votes, aes(x = imdb_votes, fill = State)) +
  geom_histogram(bins = 30, color = "white") +
  scale_x_log10(labels = function(x) format(x, scientific = FALSE, big.mark = ",")) +
  facet_wrap(~State) +
  theme_minimal()

```