---
title: "Netflix vs IMDb Quality Analysis"
author: "Meghana Mandadi, Amrita Thapa, and Thomas Zeller"
date: "4/23/2026"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(knitr)
```

```{r load-data, include=FALSE}
# --- Data cleaning setup (from Data_Cleaning_Assignment.Rmd) ---
# Produces `merged_data` used by all analysis sections below.

# Read raw data
netflix_data <- read.csv("Data/netflix_titles.csv")
imdb_data <- read.csv("Data/TMDB_all_movies.csv")

# Title cleaning
netflix_clean <- netflix_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

imdb_clean <- imdb_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

# Release year cleaning
imdb_clean <- imdb_clean %>%
  mutate(release_year = as.numeric(substr(release_date, 1, 4))) %>%
  filter(!is.na(release_year),
         release_year >= 1900,
         release_year <= 2021)

# IMDb rating and votes cleaning
imdb_clean <- imdb_clean %>%
  filter(!is.na(imdb_rating),
         !is.na(imdb_votes),
         imdb_votes > 50)

# Merge datasets
netflix_unique <- netflix_clean %>% distinct(title_clean, release_year, .keep_all = TRUE)
imdb_unique    <- imdb_clean    %>% distinct(title_clean, release_year, .keep_all = TRUE)

merged_data <- inner_join(netflix_unique, imdb_unique, by = c("title_clean", "release_year"))
```

---

# 1. Macro Trends: Quantity vs. Quality Over Time

## 1.1 How does the total volume of content added to the platform in a specific year correlate with the average IMDb score of that year's cohort?

*Visualization: Dual-Axis Line Chart or Scatter Plot with Trend Line — each point represents a year, showing the relationship between volume of content added and average IMDb rating for that cohort.*

```{r q1-volume-vs-quality}

```

## 1.2 Has the average IMDb rating of Netflix's content increased or decreased over the last decade as their production volume rapidly expanded?

*Visualization: Time Series Line Chart — `geom_line()` + `geom_smooth(method = "lm")` to show the overall slope of IMDb ratings by year.*

```{r q2-rating-trend-over-time}

```

---

# 2. Format & Genre: What Is Being Produced and How Does Quality Vary?

## 2.1 Is there a statistically significant difference in average IMDb ratings between Netflix movies and Netflix TV shows?

*Visualization: Box Plot or Violin Plot — `geom_boxplot(aes(x = type, y = imdb_rating))` to show median and spread (IQR) by content type.*

```{r q3-movies-vs-shows}

```

## 2.2 Which specific content categories (e.g., Documentaries, Stand-Up Comedy, Action) maintain the highest average IMDb ratings on the platform?

*Visualization: Reordered Horizontal Bar Chart — `geom_col()` with `fct_reorder()` to sort genres from highest to lowest average rating.*

```{r q4-ratings-by-genre}

```

## 2.3 Is there a correlation between a movie's runtime (duration) and its overall IMDb score?

*Visualization: Scatter Plot with Trend Line — each point is a movie, with runtime on the x-axis and IMDb rating on the y-axis.*

```{r q5-runtime-vs-rating}

```

---

# 3. Strategic Factors & Specific Predictors

## 3.1 Do international productions (content produced outside the United States) generally score higher on IMDb than US-based productions?

*Visualization: Box Plot or Bar Chart — compare average IMDb ratings between US and non-US productions.*

```{r q6-international-vs-domestic}

```

## 3.2 Which top five producing countries have the highest average IMDb ratings for their exported Netflix content?

*Visualization: Horizontal Bar Chart — top 5 countries ranked by average IMDb rating, filtered to countries with a minimum number of titles.*

```{r q7-top-countries-by-rating}

```

## 3.3 Does the maturity rating (e.g., TV-MA vs. TV-PG) have a strong correlation with the overall IMDb score or the total number of IMDb votes?

*Visualization: Box Plot or Grouped Bar Chart — IMDb rating and/or vote count broken down by maturity rating category.*

```{r q8-maturity-rating-vs-score}

```

## 3.4 Do shows or movies with a larger ensemble cast (based on the number of actors listed) tend to receive a higher volume of IMDb votes?

*Visualization: Scatter Plot with Trend Line — cast size on the x-axis, IMDb votes on the y-axis (log scale recommended).*

```{r q9-cast-size-vs-votes}

```

## 3.5 Do older, licensed movies (titles added to Netflix years after release) perform better in IMDb ratings than brand-new, same-year releases?

*Visualization: Box Plot or Density Plot — compare IMDb ratings between same-year releases and titles added significantly after their release year.*

```{r q10-licensed-vs-originals}

```
