---
title: "Netflix vs IMDb Quality Analysis"
author: "Meghana Mandadi, Amrita Thapa, and Thomas Zeller"
date: "4/23/2026"
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(stringr)
library(knitr)
# theme colors: olivedrab, sienna
```

```{r load-data, include=FALSE}
# --- Data cleaning setup (from Data_Cleaning_Assignment.Rmd) ---
# Produces `merged_data` used by all analysis sections below.

# Read raw data
netflix_data <- read.csv("Data/netflix_titles.csv")
imdb_data <- read.csv("Data/TMDB_all_movies.csv")

# Title cleaning
netflix_clean <- netflix_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

imdb_clean <- imdb_data %>% 
  mutate(title_clean = str_to_lower(title) %>%
         str_trim() %>%
         str_replace_all("[^a-z0-9 ]", "") %>%
         str_squish())

# Release year cleaning
imdb_clean <- imdb_clean %>%
  mutate(release_year = as.numeric(substr(release_date, 1, 4))) %>%
  filter(!is.na(release_year),
         release_year >= 1900,
         release_year <= 2021)

# IMDb rating and votes cleaning
imdb_clean <- imdb_clean %>%
  filter(!is.na(imdb_rating),
         !is.na(imdb_votes),
         imdb_votes > 50)

# Merge datasets
netflix_unique <- netflix_clean %>% distinct(title_clean, release_year, .keep_all = TRUE)
imdb_unique    <- imdb_clean    %>% distinct(title_clean, release_year, .keep_all = TRUE)

merged_data <- inner_join(netflix_unique, imdb_unique, by = c("title_clean", "release_year"))
```

---

# 1. Macro Trends: Quantity vs. Quality Over Time

## 1.1 How does the total volume of content added to the platform in a specific year correlate with the average IMDb score of that year's cohort?

*Visualization: Dual-Axis Line Chart or Scatter Plot with Trend Line — each point represents a year, showing the relationship between volume of content added and average IMDb rating for that cohort.*

```{r q1-volume-vs-quality}

colnames(merged_data)
sum(is.na(merged_data$imdb_rating))
range(merged_data$release_year)

# format year
merged_data$year_added <- as.numeric(format(as.Date(merged_data$date_added, "%B %d, %Y"), "%Y"))

# aggregate volume and average score per 'year added' cohort
analysis_data <- merged_data %>%
  filter(year_added >= 2014) %>%  #ADJUST THIS
  group_by(year_added) %>%
  summarize(
    content_volume = n(),
    avg_imdb_score = mean(imdb_rating)
  )

# Create Dual-Axis Line Chart
scale_factor <- max(analysis_data$content_volume) / 10

ggplot(analysis_data, aes(x = year_added)) +
  # Bars for volume
  geom_col(aes(y = content_volume, fill = "Content Volume"), alpha = 0.4) +
  
  # Line and Points for rating
  geom_line(aes(y = avg_imdb_score * scale_factor, color = "Avg IMDb Score"), size = 1) +
  geom_point(aes(y = avg_imdb_score * scale_factor, color = "Avg IMDb Score")) +
  
  # axis
  scale_y_continuous(
    name = "Total Content Volume",
    sec.axis = sec_axis(~./scale_factor, name = "Average IMDb Score")
  ) +
  
  # color setting
  scale_fill_manual(values = c("Content Volume" = "olivedrab")) +
  scale_color_manual(values = c("Avg IMDb Score" = "sienna")) +
  labs(
    title = "Platform Growth vs. Content Quality (by Year Added)",
    x = "Year Added to Platform",
    fill = "Metric", color = "Metric"
  ) +
  theme_minimal()


```

## 1.2 Has the average IMDb rating of Netflix's content increased or decreased over the last decade as their production volume rapidly expanded?

*Visualization: Time Series Line Chart — `geom_line()` + `geom_smooth(method = "lm")` to show the overall slope of IMDb ratings by year.*

```{r q2-rating-trend-over-time}

# Extract year and filter for years >= 2012
trend_analysis <- merged_data %>%
  mutate(year_added = as.numeric(format(as.Date(date_added, "%B %d, %Y"), "%Y"))) %>%
  filter(year_added >= 2012) %>%
  group_by(year_added) %>%
  summarize(
    avg_imdb_score = mean(imdb_rating)
  )

# Time series line chart with trend line

ggplot(trend_analysis, aes(x = year_added, y = avg_imdb_score)) +
  # Actual yearly averages
  
  geom_line(color = "olivedrab", size = 1) +
  geom_point(color = "olivedrab") +
  # Statistical trend line (Linear Model)
  geom_smooth(method = "lm", color = "sienna", linetype = "dashed", se = FALSE) +

  scale_x_continuous(breaks = seq(2012, 2022, 1)) +
  labs(
    title = "Average IMDb Rating Trend (Last Decade)",
    subtitle = "Red dashed line indicates the overall linear slope of quality",
    x = "Year Added to Netflix",
    y = "Average IMDb Score"
  ) +
  theme_minimal()

```

---

# 2. Format & Genre: What Is Being Produced and How Does Quality Vary?

## 2.1 Is there a statistically significant difference in average IMDb ratings between Netflix movies and Netflix TV shows?

*Visualization: Box Plot or Violin Plot — `geom_boxplot(aes(x = type, y = imdb_rating))` to show median and spread (IQR) by content type.*

```{r q3-movies-vs-shows}

decade_analysis <- merged_data %>%
  mutate(year_added = as.numeric(format(as.Date(date_added, "%B %d, %Y"), "%Y"))) %>%
  filter(year_added >= 2012) %>%
  group_by(year_added) %>%
  summarize(
    content_volume = n(),
    avg_imdb_score = mean(imdb_rating)
  )

# bar chart with trend line

ggplot(decade_analysis, aes(x = year_added, y = content_volume)) +
  geom_col(fill = "olivedrab", alpha = 0.6) +
  # Add the linear trend line to show the rate of expansion
  geom_smooth(method = "lm", color = "sienna", se = FALSE, linetype = "solid") +
  scale_x_continuous(breaks = seq(2012, 2022, 1)) +
  labs(
    title = "Netflix Production Volume Expansion (2012-2022)",
    subtitle = "Solid red line shows the linear growth trajectory of content library",
    x = "Year Added",
    y = "Number of Titles Added"
  ) +
  theme_minimal()

```

## 2.2 Which specific content categories (e.g., Documentaries, Stand-Up Comedy, Action) maintain the highest average IMDb ratings on the platform?
```{r q4-ratings-by-genre}
# Split genres
genre_split <- strsplit(as.character(merged_data$listed_in), ",")

# Expand into long format
genre_data <- data.frame(
  genre = trimws(unlist(genre_split)),
  imdb_rating = rep(merged_data$imdb_rating, lengths(genre_split))
)

# Remove missing ratings
genre_data <- genre_data[!is.na(genre_data$imdb_rating), ]

# Summarize
genre_summary <- genre_data %>%
  group_by(genre) %>%
  summarize(
    avg_rating = mean(imdb_rating, na.rm = TRUE),
    count = n()
  ) %>%
  filter(count >= 30) %>%
  arrange(desc(avg_rating))

ggplot(genre_summary, aes(x = reorder(genre, avg_rating), y = avg_rating)) +
  geom_col(fill = "olivedrab", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Average IMDb Ratings by Genre (n ≥ 30)",
    x = "Genre",
    y = "Average IMDb Rating"
  ) +
  theme_minimal()
```

## 2.3 Is there a correlation between a movie's runtime (duration) and its overall IMDb score?
```{r q5-runtime-vs-rating}
# clean runtime movies only
movie_data <- merged_data %>%
  filter(type == "Movie") %>%
  mutate(
    runtime = as.numeric(str_extract(duration, "\\d+"))
  ) %>%
  filter(!is.na(runtime), !is.na(imdb_rating))

cor(movie_data$runtime, movie_data$imdb_rating)

ggplot(movie_data, aes(x = runtime, y = imdb_rating)) +
  geom_point(alpha = 0.3, color = "olivedrab") +
  geom_smooth(method = "lm", color = "sienna", se = FALSE) +
  labs(
    title = "Relationship Between Movie Runtime and IMDb Rating",
    x = "Runtime (minutes)",
    y = "IMDb Rating"
  ) +
  theme_minimal()
```

---

# 3. Strategic Factors & Specific Predictors

## 3.1 Do international productions (content produced outside the United States) generally score higher on IMDb than US-based productions?
```{r q6-international-vs-domestic}
# us vs international variable
country_data <- merged_data %>%
  filter(!is.na(country), !is.na(imdb_rating)) %>%
  mutate(
    origin = ifelse(str_detect(country, "United States"),
                    "US Production",
                    "International Production")
  )

# averages
country_data %>%
  group_by(origin) %>%
  summarize(
    avg_rating = mean(imdb_rating, na.rm = TRUE),
    count = n()
  )

t.test(imdb_rating ~ origin, data = country_data)

ggplot(country_data, aes(x = origin, y = imdb_rating, fill = origin)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = c("US Production" = "olivedrab",
                               "International Production" = "sienna")) +
  labs(
    title = "IMDb Ratings: US vs International Productions",
    x = "Production Origin",
    y = "IMDb Rating"
  ) +
  theme_minimal()
```

## 3.2 Which top five producing countries have the highest average IMDb ratings for their exported Netflix content?

*Visualization: Horizontal Bar Chart — top 5 countries ranked by average IMDb rating, filtered to countries with a minimum number of titles.*

```{r q7-top-countries-by-rating}

```

## 3.3 Does the maturity rating (e.g., TV-MA vs. TV-PG) have a strong correlation with the overall IMDb score or the total number of IMDb votes?

*Visualization: Box Plot or Grouped Bar Chart — IMDb rating and/or vote count broken down by maturity rating category.*

```{r q8-maturity-rating-vs-score}

```

## 3.4 Do shows or movies with a larger ensemble cast (based on the number of actors listed) tend to receive a higher volume of IMDb votes?

*Visualization: Scatter Plot with Trend Line — cast size on the x-axis, IMDb votes on the y-axis (log scale recommended).*

```{r q9-cast-size-vs-votes}

```

## 3.5 Do older, licensed movies (titles added to Netflix years after release) perform better in IMDb ratings than brand-new, same-year releases?

*Visualization: Box Plot or Density Plot — compare IMDb ratings between same-year releases and titles added significantly after their release year.*

```{r q10-licensed-vs-originals}

```
